Omics-based Single Step Trait Prediction

### 2017-09-06 paper/manuscript.tex
paper, describe new subsampling scheme

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
The maize hybrid data set, subsequently denoted as \textit{Exp1}, comprised 
1,521 hybrids produced in 16 factorial mating designs between 142 Dent and 103 
Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
gene expression BLUEs were computed for 1,323 transcripts while adjusting for 
putative effects arising from the presence of multiple factorials (\textit{cf.}
\citeNP{Westhues2017}) using the R-package \textit{limma}~\cite{Ritchie2015a}.

This dataset was included to address objectives 1 and 2.

\Genetics2level{Maize diversity panel}
The maize inbred line data set, subsequently referred to as \textit{Exp2}, 
originally comprised 513 maize lines representing the global maize diversity 
and was reduced to the set of tropical and subtropical lines ($n = 211$) given 
that it is the largest of the four pre-classified subgroups~\cite{Yang2014}.
We reduced the original dataset to this subset to avoid inflated predictive 
abilities due to large differences in the average performance of the different
subpopulations~\cite{Windhausen2012}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

This dataset was included to address objectives 2 and 3.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}lll@{}}
\toprule
Trait              & Acronym & Data Set  \\ \midrule
Cob weight         & CW      & Diversity \\
Days to silking    & DS      & Diversity \\
Ear diameter       & ED      & Diversity \\
100-grain weight   & GW      & Diversity \\
Kernel width       & KW      & Diversity \\
Plant height       & PH      & Diversity \\
Dry matter yield   & DMY     & Hybrid    \\
Dry matter content & DMC     & Hybrid    \\
Fat                & FAT     & Hybrid    \\
Protein            & PRO     & Hybrid    \\
Starch             & STA     & Hybrid    \\
Sugar              & SUG     & Hybrid    \\ \bottomrule
\end{tabular}
\end{table}

\Genetics2level{Population structure}
Separately for each data set, principal component analyses (PCA) were run on 
scaled and centered predictor matrices using the \emph{pca} function from the 
R-package \textit{LEA}~\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Algorithms were run for $K=2, \dots 5$ putative ancestral gene pools with 25
repetitions for each value of $K$. 
For each $K$, only the repetition with the lowest cross-entropy was kept for 
further analyses.

\Genetics2level{Core sampling}
In order to evaluate the influence of the genetic constitution of the set of
genotypes that has only data on one out of two predictors, we generated core
samples based on two different approaches.
All core sets were assembled from the 149 maize inbred lines from the diversity
panel which had gene expression as well as genomic information.
We defined core samples as a subset of genotypes that is covered by two
predictors whereas all other genotypes (\textit{i.e.}, the complement) are only
covered by a single predictor.
The first application involved assembling core sets of varying sizes by 
maximizing the average genetic distance among core set members using the 
Modified Rogers (MR) distance as the criterion~\cite{Thachuk2009}.
The size of the core set was varied in increments of ten percentage points and 
ranged from 10\% to 90\% of all 149 inbred lines.
For the second application, we first ran an admixture analysis in combination
with a PCA and kept $K=3$ as the number of putative ancestral population, which 
seemed to describe the population structure best.
Genotypes were then assigned to one of the three ancestral populations based on
whether any of the three relationship coefficients (A1-A3) was equal to or
greater than 0.5.
For example, a genotype with an admixture coefficient for A2 $\geq 0.5$ was
assigned to the second ancestral population.
Thereby, we obtained $n_{A1} = 17$, $n_{A2} = 109$, $n_{A3} = 19$ and four 
genotypes that could not be unambiguously assigned to any putative ancestral 
population based on this criterion.
To ensure commensurability of the three subpopulations within the set of 149
genotypes, the subset of 109 genotypes (\textit{i.e.}, members of A2) was
reduced to $n_{A2} = 19$ genotypes, again using the method of maximizing the
average genetic distance among selected genotypes described above.

@@

>>>>

\section{Materials and Methods}
\Genetics2level{Maize hybrids}
To address the objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp1}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

\Genetics2level{Maize diversity panel}
To address objectives 2 and 3 a maize inbred line data set, subsequently
referred to as \textit{Exp2}, was used~\cite{Yang2014}.
Exp2 originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}lll@{}}
\toprule
Trait              & Abbreviation & Data Set  \\ \midrule
Cob weight         & CW           & Diversity \\
Days to silking    & DS           & Diversity \\
Ear diameter       & ED           & Diversity \\
100-grain weight   & GW           & Diversity \\
Kernel width       & KW           & Diversity \\
Plant height       & PH           & Diversity \\
Dry matter yield   & DMY          & Hybrid    \\
Dry matter content & DMC          & Hybrid    \\
Fat                & FAT          & Hybrid    \\
Protein            & PRO          & Hybrid    \\
Starch             & STA          & Hybrid    \\
Sugar              & SUG          & Hybrid    \\ \bottomrule
\end{tabular}
\end{table}

\Genetics2level{Population structure}
Population structure in Exp2 was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:pca}.

@@

\Genetics2level{Core sampling}
In order to evaluate the influence of the fraction of genotypes that is not
covered by the incomplete predictor on predictive ability we generated core
sets of nine different sizes from Exp2 (Fig.~\ref{fig:compound-plot}~a).
The base of each of these core sets was the superset of genotypes that were
covered by both, genomic and trancriptomic data.
A core set was created by artificially declaring a pre-specified fraction 
$x \in {10\%, 20\%, \dots, 90\%}$ of genotypes as lacking of transcriptomic 
information.
For each value of $x$ we created 100 core sets $s \in {1, 2, 3, \dots, 100}$ by
repeatedly sampling genotypes, whose trancsriptomic information was then 
declared as missing, at random (Fig.~\ref{fig:compound-plot}~b).
The artificially removed transriptomic information on these genotypes was then
imputed.
